Writing Trash Bin

### 2024-03-09 paper.qmd
finished definitions in existing work section

> Edit note: just summarise common themes of issues the papers ran into instead of detailling the work paper by paper.

## Types of Uncertainty Visualisation Papers

# The Uncertainty Definition Problem
## Definitions in Existing Work
Uncertainty visualisation is not made any easier by the fact that the term "uncertainty" lacks a commonly accepted definition in the literature. @Spiegelhalter2017 even commented that the definition of uncertainty essentially "depends on who you ask". The disarray of organisation and definitions within the uncertainty literature is a common sentiment expressed by reviewers[@Spiegelhalter2017; @Kinkeldey2014]. This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is. In this section we will establish exactly what we mean when we talk about uncertainty.

The most common definition of uncertainty used in papers that discuss visualising it are colloquial. Most definitions make some nod towards variance, precision, trust, error, missing values or another related concepts. Since *uncertainty*, unlike *probability*, does not have a strict definition that is established in mathematics, there is large variety in its usage. Some consider it to be synonymous with probabiltiy, some consider it to be an encompassing umbrella term of which probability is only an element, and some consider it to by something else entirely. This lack of a definition leads to uncertainty being swapped out freely with other vague words such as ambiguity and risk. Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty". The issue with these broad definitions is is that in order to visualize something, it needs to be quantified, and in order to quantify something, we need to know exactly what it is we are counting. This discrepancy appears regularly in the uncertainty visualisation literature. Often a vague definition such as "uncertainty is anything that isn't deterministic", or even a refusal to define uncertainty at all in an experiments introduction will be accompanied by a highly specific quantification of uncertainty as a probability density distribution of a forecast when providing the visualisation. Ironically it seems that many of the papers that refer to themselves as "uncertainty" papers are just suffering from a form of uncertainty, a lack of precision.[^1]

This lack of precision has caused several pernicious issues in the literature, including different representation of uncertainty being compared despite not being realistic substitutes (with no acknowledgement of this from the authors); a lack of understanding as to *why* uncertainty should be included in a visualisation; a large swath of literature focused on a minute aspect of the entire uncertainty space with little attention to creative solutions to real world visualisation problems; and a swath of newly designed visualisations that are disconnected from the problems that they hope to solve. These problems means the literature either needs to transition away from the broad umbrella term "uncertainty", or consider the full range of sources of uncertainty and how they can be communicated when designing an experiment.[^2]

The most encompassing definition of uncertainty I have seen comes from @utypo who define uncertainty as **"any deviation from the unachievable ideal of completely deterministic knowledge of the relevant system"**. This definition encapsulates many concepts adjacent to randomness such as predictions, probability mass distributions (PMF), estimate error, and any data that is not a set of deterministic outcomes. This is the definition of uncertainty I will use for the rest of this report.

[^1]: Flesh out note - I think several of the sentences in this paragraph need examples (or quatified evidence) from the spreadsheet definition part
[^2]: I am not sure if I should mention this here because I go into depth in the literature review and these comments might come across as unjustified if they are left in this section. Maybe this is more of a conclusion paragraph.

### Taxonomy Definitions
> Edit Note: A lot of this is going to be moved to the uncertainty definition specific section of the vis lit review and the point here should just be "this is a mess" instead of trying to find any reason in it. My opinion and goal for this section has changed a lot since writing this (specifically deciding that the encapsulating taxonomy doesnt fix the problem and a maths definition is needed) so it is one of the sections that needs big rewrites.

While colloquial definitions of uncertainty are useful when trying to communicate with laypeople who don't understand strict statistical definitions, these definitions are not helpful when we are trying to work out exactly what should be shown on a plot. When authors try to make a considerable effort to define what we mean when we talk about uncertainty, they often express it in the form of a taxonomy.

When reading any literature about uncertainty or an adjacent topic you will be overrun with more ways to organise uncertainty than you will know what to do with. The problem is so pervasive that it is almost impossible for two randomly selected papers that both claim to be about uncertainty will use the same definition. What is worse is that the reason for using one taxonomy over another, or why a new taxonomy was established in the first place is rarely discussed. This leaves new authors in the field with a wide array of choices and no information that would allow them to determine which is most effective for them. The complete mess of this literature makes it more enticing for authors to just define their own definition of uncertainty, which only adds to this problem. This is not done in ignorance of the issue either, many papers that comment on the unending list of uncertainty definitions will still provide a new taxonomy of their own design. I would almost find this situation comical if I were not one of the people forced to sort through this mess.

Examples of Taxonomies used to untanlge the features of uncertainty:

- aleatory uncertainty (unavoidable randomness), epistemic uncertatiny (uncertainty about structure, still quantified), ontological uncertainty (uncertain about entire modelling process). @Spiegelhalter2017
- measurement (uncertainty on a single varible), spatial (uncertainty in spatial data), temporal (uncertainty in a time data)
- aleatory (inherrent randomness in a forecast) vs epistemic (uncertainty in structure and parameters of statistical models) vs ontological (uncertainty about the entire modelling process as a description of reality)
- scientific methods (variance, distribution) vs human judgments (disagreement among parties)
- Random vs systematic uncertainty
- statistical (can provide a continuous pdf) vs bounded (know values will fall in a certain range) uncertainty
- accuracy vs precision
- Uncertainty from discovery, assessment and execution
- risk vs uncertainty
- "known knowns, known unknowns, and unknown unknowns", (risk, uncertainty, ignorance)
- Source of uncertainty (model structure, parameters, forcing functions, intial state, model operation) and diagnostic (past/current) vs prognostic (forecasting)
- uncertainty is organised according to Satience (legal moral, societal, institutional, proprietary, situational) and level of severity (high, medium, low)
- Level of precision in epistemic uncertainty: 1) numbers given to appropriate levels of precision 2) a distribution range 3) a measure of statistical significance 4) verbal quantifiers to numbers 5) refusal to give a number unless the evidence is good enough @Spiegelhalter2017

![Example of different uncertainty taxonomies](paperscreenshots/collage1.jpg){#fig-location}

> Edit note: I dont think I will keep this figure in the paper. The goal was to show they are inconsistent but I think a sentance or two in the text will achieve that.

This taxonomy issue creates several problems in the literature. Ideally, if we wanted to combine the literature on uncertainty visualisation, each paper would state the same definition of uncertainty and explain which aspect of uncertainty they are visualising, so the experiments would shift around the space while the definition of uncertainty and what it contains would remains stationary. This would make it easy for the typical reader to use the literature to find an appropriate visualisation for the particular category or case of uncertainty. Instead, one has to comb through the method of a particular design to understand if the uncertainty you are working with is amenable to this method. Another issue this creates is that the "space" of uncertainty visualisation is unclear. This means that several areas of uncertainty visualisation that have a real need for creative visualisation tools are left untouched, while the visualisations that are considered "uncertainty visualisations" simply because they are an expression of a probability mass function dominate the literature. Finally the taxonomies do little to help data analysts or statisticians untangle, estimate and understand the uncertainty that may be in their own projects. The endless set of constantly changing definitions implies their *is no* set meaning to the term "uncertainty" and something that cannot be defined certainly cannot be quantified for communication or estimation. This leads people to be less likely to visualise uncertainty due to its imprecision and difficulty to commincate.

This does not mean there is nothing to be learned from these taxonomies, each in their own way show what apsect of uncertainty is important to certain cases. There are a handful of overlapping features that indicate which elements of uncertainty are of interest to statistics and related applied fields. @utypo identified these similarities and used them to design a taxonomy that encapsulates a large proportion of the ad hoc definitions that currently overrun the literature, and is the definition of uncertainty that we will work with for the remainder of this paper.

@fig-taxonomy is an illustration of the taxonomy presented by @utypo. In this taxonomy, there are three things we need to consider for each "uncertainty" we encounter through the modelling process. First, we need to consider the source of the uncertainty. Is this uncertainty coming from the data (for example, from an inaccurate measurement), from the array of model choices we have, or from the assumptions on a parameter? This is the *location* of the uncertainty. Second, consider how well you can quantify this uncertainty. Do you know exactly how much measurement error there is in each observation or are you not even aware if there is a measurement error? This is the *level* of your uncertainty, and it ranges from discrete to total ignorance. Finally, consider how this uncertainty came into existence. Is it a result of a naturally random process (aleatory) or is it due to imperfect information and could be improved (epistemic). This is the *nature* of your uncertainty. @utypo then goes on to describe mapping our uncertainty in a 3D space that is defined by its location, level, and nature, but I think the taxonomy is more easily understood as a series of questions we should to consider when we are trying to quantify uncertainty. Location asks "where and at what stage in the modelling process is this uncertainty coming from", level asks "how well can we quantify this uncertainty", and nature asks "can we reduce this uncertainty with improves knowledge?". While this taxonomy specifically applies to the uncertainty in the model, it is easy to fit most of the uncertainty we would consider quantifying into this format.

@@

> Edit note: Everything from this point down was used to try and provide some justification for why this dimensionality concept could work, but I have since changed my mind and am probably going to get rid of this section unless it fits somewhere else.

*The taxonomy* we will use in this paper may seem complicated and difficult to grasp at first glance, but it essentially boils down to three dimensions to consider. Uncertainty is a broad and complicated topic and while oversimplified taxonomies might help us hold the concept of uncertainty in our head, they do little to help with the practical question of how to quantify, visualise, or communicate uncertainty. Organising uncertainty according to its location, level and nature makes allows us to consider uncertainty in the miriad of ways it can arise in a data analysis project and better understand *how* this uncertainty should be communicated. With a detailed understanding of the sources of uncertainty and the ways it can be quantified, we can better understand the gaps in the current set of visualisation tools and assess if these gaps need to be filled.

*The location* axis of the taxonomy lines up neatly with the typical data analysis pipeline. Multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Hullman2016; @Refsgaard2007]. @fig-location shows the connection between the typical data analysis pipeline and the location element of the uncertainty taxonomy, along with examples of how uncertainty could arise from each of these steps. The relevance of every stage of the uncertainty taxonomy is backed up by both theoretical and practical work in statistics and adjacent fields. @Munzner2009, in her nested model for visualisation, comments on how even something as low level as poor problem abstraction can cause down stream effects and become impossible to ignore. @Otsuka2023 highlights that the choices made in the ontological process of boiling real world events down into statistical objects can influence later stages of analysis due to what was considred signal and what was considered noise. @Meng2021 stresses the importance of treating data as an input from another process with noise and variance, rather than as a neutral input. High level stages such as model selection and estimation have obvious introductions of uncertainty through the choice of models and nature of estimation [^I'm not sure if I need a citation here because this feels obvious]. Location is also acknowledged as an important element of consideration in the taxonomies of several visualsiation papers [^ I will probably need a citation for which ones].

![Illustration of the connection between the data generating process and the location of the uncertainty analysis. Several examples are collected from utypo and Munzner2009. ](location.png){#fig-location}

*The level* axis forces us to identify how quantifiable the uncertainty is. This element is important because establishing how quantifiable uncertainty is informs us how it can be communicated, something that @Spiegelhalter2017 also identified in the form of "precision". Statistical uncertainty can be quantified, so it can be expressed using a PDF, a variance, an error estimate, or something similar while outcome uncertainty cant be quantified but each scenario may be simulated. Acknowledged ignorance can only be stated in the form of an assumption. Visualisations of climate change scenario uncertainty typically combine the outcome uncertainty of carbon emissions with the and statistical uncertainty of the noisy measurement and modelling system to create a cohesive illustration of the uncertainty surrounding climate change. It is commonly noted that each uncertainty must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017]. The level axis and climate change visualisations provide tools for how to consider combining these uncertainties. Uncertainties that are the same level but from multiple locations can likely be combined to establish an overall model uncertainty. While uncertainties at different levels cannot be combined, they can still be visualised simultaneously as shown in the climate scenario uncertainty visualisations.

*The nature axis* allows us to easily communicate if the uncertainty can be reduced or if it needs to be accepted. While this level if not of particular importance to the visualisation of uncertainty, that is not the only purpose of the taxonomy. The nature of the uncertainty is an important element in expressing uncertainty, especially to stakeholders, and connecting the nature to the location makes it easier to understand *how* uncertainty can be reduced.

Mapping uncertainty into the 3D space defined by location, level, and nature can expand beyond the narrow case of statistical modelling it was defined for. *(find the citation for this)* included "consensus" as a taxonomy element, however it was not included in the final definition, but it can easily be added as a location that would align with the communication stage of the data pipeline. Several authors [@Spiegelhalter2017] *(I have other papers)* considered the method by which uncertainty was quantified (e.g. by a single value, a function, simulated outcomes, etc), which can be be thought of as a more precise categorisation of the level axis. This taxonomy also ignores contextual information about the data, such as spatial or temporal considerations. This flexibility means that the use of this framework is not limited by a specific case and should be used going forward to define what aspect of the uncertainty space a contribution exists in.

It is important to note that this taxonomy is thorough but it is not a mathematical definition of uncertainty. While individual *expressions of uncertainty*, such as a confidence interval or a PDF are mathematically defined, the term *uncertainty* has not been defined mathematically. This is largely because it is a layperson word that has been adopted by statistics to describe any deviation from complete determinism, however this does not mean that a mathematical definition that encompases missing values, alternative models, and prediction intervals *cannot* be found. However, that task is far beyond the scope of this work. In this work, we will investigate the array of misunderstandings and focus on inspecting how the current literature *deviates* from already established mathematical definitions, why these deviations occur, and understand how these deviations can lead to larger problems in uncertainty communication.

@@

# The Uncertainty Visualisation Definition Problem

@@

## "Uncertainty Visualisation" is not defined

@@

# Measureing noise, not signal

>>>>

# The Definition Problem
## Definitions in Existing Work
Uncertainty visualisation is made particularly difficult by the term "uncertainty" lacking a commonly accepted definition in the literature. This mishmash of terminology leads to a large body of work, all claiming to finding the best visualisation or expression of of "uncertainty" but most don't even seem to agree on what uncertainty is.

Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty". Some consider it to be synonymous with specific terms defined in mathematics, such as probability, variance, error, or precision. Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples, and include concepts that are somewhat related, such as missing values (*Cite Gap 3*). This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[], ignore defining uncertainty entirely [], or pick a definition from a seemingly randomly previously published papers [], all of which make variance in uncertainty definitions even worse (*Cite Gap 3.5*).

While these vague definitions might be OK for polite conversation, they are *not* foundation on which to build an entire sub-field of visualisation. In order to visualise uncertainty, it needs to be quantifiable [@Griethe2006; Leland2005], in order to quantify uncertainty, it needs to be mathematically defined. The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase, (e.g. @utypo defined it as any deviation from complete determinism) but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation (*Cite Gap 4*). This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods dont even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [], mass [], or variance [] (*Cite Gap 5*). This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention (*Cite Gap 6*). There is likely only confusion and obfuscation if the vague definitions of "uncertainty" continue to be used and authors should try to be more specific with the mathematical object they are visualising, *or* a strict definition of uncertainty needs to be defined. The current method of titling papers and defining uncertainty forces readers to comb through the method of a particular visualisation experiment to understand if the uncertainty visualised is amenable to their needs.

Some papers have recognised these colloquial definitions are lacking and have made attempts to formalise the space with taxonomies, however these taxonomies come with their own issues.

### Taxonomy Definitions
Much in the same way that there are almost as many definitions of uncertainty as there are papers on the topic, the field is also over run with taxonomies. Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], etc., etc., etc. (*Cite Gap 7*). In their own way, each of these taxonomies show an apsect of uncertainty that an author felt was important to differentiatet. @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy, depicted in @fig-taxonomy. This definition organises uncertainty based on three axis, location (where the uncertianty is coming from), level (how quantifiable it is), and nature (if it comes from true randomness or incomplete knowledge).

@@

When considering the use of taxonomies to define something inherrently vague, such as the uncertainty in an uncertainty visualisation, it is helpful to consider this quote from *The Grammar of Graphics*:

> "Taxonomies are useful to scientists when they lead to new theory or stimulate insights into a problem that previous theorizing might conceal. Classification for its own sake, however, is as unproductive in design as it is in science. [@Leland2005]

Therefore, while taxonomies can be helpful to map out the space of "things that we consider to be related to uncertainty", a taxonomy is *not* a definition and do little to help statisticians untangle and estimate the uncertainty in their projects. That being said, these taxonomies can hint towards what is important when we think about uncertainty. The location axis of the @utypo taxonomy lines up neatly with the typical data analysis pipeline and multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Hullman2016; @Refsgaard2007] including stages as early as conceptualising the problem [@Otsuka2023] or collecting the data [@Meng2021]. The level axis from the @utypo taxonomy also hints towards an important consideration in uncertainty, because establishing how quantifiable uncertainty is informs us how it can be communicated, something that @Spiegelhalter2017 identified in the form of "precision". It is commonly noted that each source of uncertainty in an analysis must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017] (*Cite Gap 8*). It is clear that elements of these taxonomies are identified in many other comments on uncertainty and represent *real* considerations that need to be made when defining uncertainty, even if that definition is yet to be published.

The task of providing an encapsulating mathematical definition of uncertainty is far beyond the scope of this work, however we will discuss an important but commonly misunderstood feature of uncertainty; its relationship to inference.

@@

# The Visualisation Problem

@@

### Types of Uncertainty Visualisation Papers

## Uncertainty Visualisation is not defined

@@

# The Measureing Noise Problem

### 2024-03-09 writingdumpster.qmd
finished definitions in existing work section

@@

>>>>

Cite Gap 3: Some consider it to be synonymous with specific terms defined in mathematics, such as probability, variance, error, or precision. Others consider it to be an encompassing umbrella term of which for these mathematically defined objects are only examples, and include concepts that are somewhat related, such as missing values

Cite Gap 3.5: This wide array of uncertainty definitions becomes a snake eating its own tail, as many authors opt to redefine uncertainty themselves[], ignore defining uncertainty entirely [], or pick a definition from a seemingly randomly previously published papers [], all of which make variance in uncertainty definitions even worse

Cite Gap 4:The uncertainty visualisation literature is completely awash with papers that define uncertainty using a vague encompassing phrase but go on to *quantify* uncertainty as a PDF, error, or some other easily quantified mathematical object when it comes time to do the visualisation.

Cite Gap 5: This definition swap happens so subtly that it seems to go unnoticed by authors writing the papers, and the fact that "uncertainty visualisation" methods dont even notice they are exclusively focus on depictions of easily quantifiable uncertainty, such as error [], mass [], or variance []

Cite Gap 6: This has left the expressions of uncertainty that are hard to quantify or visually differentiate untouched, despite many papers calling for their invention

Cite Gap 7: Taxonomies split uncertainty based on an endless stream of arbitrary lines in the sand, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017], if the uncertainty is in the the attribute, spatial elements, or temporal element of the data [], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [], if the uncertainty is random or systematic [], statistical or bounded [], accuracy or precision [], if the uncertainty is about the past or the future [], the stage of the data analysis pipeline it comes from [], the severity of the uncertainty [], how quantifiable the uncertainty is [@Spiegelhalter2017], and many, many, many more examples.

Cite Gap 8: It is commonly noted that each uncertainty must be discussed in isolation, but combining the uncertainty from every stage is near impossible [@Spiegelhalter2017]

@@

- @Munzner2009, in her nested model for visualisation, comments on how even something as low level as poor problem abstraction can cause down stream effects and become impossible to ignore. 
- @Otsuka2023 highlights that the choices made in the ontological process of boiling real world events down into statistical objects can influence later stages of analysis due to what was considred signal and what was considered noise. 
- @Meng2021 stresses the importance of treating data as an input from another process with noise and variance, rather than as a neutral input.
